# Analysis

## Exploratory data analysis

```{r message=FALSE, warning=FALSE}
library(dendextend)
library(factoextra)
library(GGally)
library(ggfortify)
library(ggrepel)
library(gridExtra)
library(knitr)
library(kableExtra)
library(mclust)
library(NbClust)
library(plyr)
library(tidyverse)
library(ggthemes)
library(tictoc)
# set for reproducible results
set.seed(14)
# set theme for ggplot
theme_set(theme_minimal())
# clear variables
rm(list = ls())

# Load custom theme
theme_davide <- function() {
  theme_fivethirtyeight(base_family = 'avenir') %+replace%  
    theme(
      text = element_text(family='Pt Mono'), 
      axis.title.x = element_text(color = 'black', margin = margin(t = 30, b = 8), family = 'K2D', face = 'bold', size = 19), 
      axis.title.y = element_text(color = 'black', margin = margin(r = 25, l = 8), family = 'K2D', face = 'bold', size = 19, angle = 90), 
      axis.text = element_text(color = 'grey30'),
      axis.text.x = element_text(face='bold', size = 13),
      axis.text.y = element_text(face='bold', size = 13), 
      panel.background = element_rect('grey98'), 
      plot.background = element_rect('grey98'),
      plot.title = element_text(margin = margin(b=15, t = 10), face='bold', size=30, hjust = 0.5, family = 'Proxima Nova'),
      plot.subtitle=element_text(size=15, hjust = 0, margin = margin(b = 10), family = 'Proxima Nova'), 
      panel.grid.major = element_line(color='gray80', linetype = 'dashed'),
      plot.margin = unit(c(0.5, 1, 0, 0.2), "inches"), 
    ) 
}
```

We decided to use the 14 features in our analysis (on a per game basis) because they provide a good balance of offensive (e.g Pts, Ast) and defensive stats (e.g. Reb, Blk). This is more likely to provide more balanced groupings between those who are more offensive and those who are better at defense.

With these features, we will explore the distribution of each future by plot the distributions of key per game stats.

```{r}
# load data
nba <- read.csv('/Users/davidetissino/Desktop/Tesi/Thesis/Datasets/Model_clean_stats.csv')

# replace NA values in RPM column with 0s
nba$BPM[is.na(nba$RPM)] <- 0
```


```{r echo=FALSE}
# plot distributions of key per game stats
# grep matches all the _pg columns into one dataframe nba_feat
features_pg <- grep('_pg', names(nba), value = TRUE)
nba_feat <- nba[ , features_pg]

# plots densities of per game stats
nba_feat %>% 
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~name, scales = "free") + 
  labs(title = 'Feature densities (un-transformed)', 
       x = '')
```

Variance was examined to provide a numerical estimate of the spread of each feature to determine if scaling is necessary. As an exploratory analysis, bi-variate comparisons were made between all features to determine if there are any groupings between variables. A subset of the features is presented below.

```{r echo=FALSE}
# calculate variance of each per game stat
var_table <- round(apply(nba_feat, MARGIN = 2, FUN = var), 2)

var_df <- data.frame(var_table)

colnames(var_df) <- 'Variance'

# plot table of variance
kable(t(var_df),
      caption = 'Feature Variance', 
      booktabs = TRUE) %>% 
  kable_styling(latex_options = c("striped", "scale_down"))
```


```{r, echo=FALSE, warning=FALSE}
# look at pairs plots for key stats
feat_plot <- c('MP_pg', 'PTS_pg', 'AST_pg', 'TRB_pg', 
               'STL_pg', 'BLK_pg')
nba_feat_plot_pos <- nba[ , feat_plot]

```


Due to the discrepancy in spread between features we decided to scale the data. Scaling each feature ensures that no single feature will dominate subsequent analyses as a result of the way the feature was measured. A good example is minutes and blocks per game - most players will have more minutes per game than blocks per game.

```{r}
# scale the per game stats
nba_feat_sc <- scale(nba_feat)
```


## Principal component analysis

Before running any clustering algorithms, we will perform principal component analysis to determine if there are any inherent groupings among players.

The first two principal components explain the majority of the variance in the feature set. We will plot the data in these two dimensions to better assess player similarities.

```{r}
# run PCA on scaled per game stats
# 
nba_pca <- prcomp(nba_feat_sc)

# each PC are the major stats
# PC1 = Minutes
# PC2 = Field Goals
summary(nba_pca)
```

```{r}
# distance matrix for features
nba_dist_sc <- dist(nba_feat_sc, method = 'euclidean')
  
# try single, centroid, and ward (D2) linkage hier clustering
hcl_single <- hclust(d = nba_dist_sc, method = 'single')
hcl_single
hcl_centroid <- hclust(d = nba_dist_sc, method = 'centroid')
hcl_centroid
hcl_ward <-  hclust(d = nba_dist_sc, method = 'ward.D2')
hcl_ward
```



```{r}
# create a function that will plot PCA
plot_pca <- function(object, frame = FALSE, x = 1, y = 2, 
                     data, colour, title, label) {#, leg_title) {
  # plots data in PCA space 
  # object = PCA or K-Means object
  # x = which PC for x-axis (1, 2, ,3, etc..)
  # y = which PC for y-axis (1, 2, 3, etc..)
  # object: PCA or K-means object
  # data = underlying data
  p <- autoplot(nba_pca, x = x, y = y, data = nba, colour = colour, frame = frame) + 
        ggtitle(title) +
        # center title
        theme(
          plot.title = element_text(hjust = 0.5)
          )# + 
        # geom_label_repel(
        #   aes(label = label),
        #   box.padding   = 0.15, 
        #   point.padding = 0.4,
        #   segment.color = 'grey50'
        #   )
  ## This is supposed to override the autoplot legend title.
    ## Only works when plotting PCA directly. Does not work for HCL and KM objects
    #labs(colour = leg_title)
  return(p)

}


```


At first glance, there are differences between players based on overall statistics. For example, LeBron James and James Harden are near each other, indicating star players may be grouped together. There are also similarities based on player position. For example, Centers/Power Forwards such as Anthony Davis and Karl-Anthony Towns are in the bottom of the chart.

Examining the PCA in three-dimensional space shows a 'cone' shape spread of the data similar to the two-dimensional  'fan' shape.

```{r echo=FALSE}
# Labels: Players who played more than 36 min per game or less than 3 min per game
labels_pca <- ifelse(nba$MP_pg >= 36 | nba$MP_pg <= 2, 
                    as.character(nba$Player), '')
title_pca <- paste0('PCA: NBA - ', ncol(nba_feat) ,' features')

# Plot first two components with positions
plot_pca(nba_pca, data = nba, colour = 'Pos', 
         label = labels_pca, title = title_pca) + 
  geom_point(
    shape = 21, 
    size = 3, 
    color = 'grey80')+ 
  theme_davide() + 
  theme(
    legend.margin = margin(4, 4, 4, 4), 
    legend.background = element_rect(fill = 'grey98'), 
    legend.position = 'bottom',
    plot.title = element_text(margin = margin(b=15, t = 10), size=30, hjust = 0.5, family = 'Proxima Nova', face = 'bold'),
    plot.margin = margin(5, 0, 5, 0)
    )#, leg_title = 'Position'


# 
ggsave('/Users/davidetissino/Desktop/PCA_features.png', dpi = 'retina', width = 9, height = 9)

```





### K-Means

#### Optimize number of clusters

Similar to the optimization for hierarchical, we found similar results for K-Means. We decided with the 4-cluster solution based on the Silhouette index (local maximum).  

```{r}
# get optimal cluster sizes 
cluster_sizes_km_ch <- NbClust(data = nba_feat_sc,
                          # it will likely be harder to interpret clusters
                          # past this amount
                          max.nc = 6,
                          method = 'kmeans',
                          index = 'ch')

# get optimal cluster sizes 
cluster_sizes_km_s <- NbClust(data = nba_feat_sc,
                          # it will likely be harder to interpret clusters
                          # past this amount
                          max.nc = 6,
                          method = 'kmeans',
                          index = 'silhouette')
```

```{r echo=FALSE}
par(mfrow = c(1,2))

# plot C(G)
plot(names(cluster_sizes_km_ch$All.index),
     cluster_sizes_km_ch$All.index,
     main = 'Calinski-Harabasz index: KM',
     type = 'l',
     xlab = "Cluster size",
     ylab = "Index value")


# plot C(G)
plot(names(cluster_sizes_km_s$All.index),
     cluster_sizes_km_s$All.index,
     main = 'Silhouette Index',
     type = 'l', 
     xlab = 'Cluster size',
     ylab = 'Index value')

#ggsave('/Users/davidetissino/Desktop/Cluster_Sizing.png', dpi = 'retina', width = 9, height = 9)

```


#### K-means clustering with four groups

Compared to the hierarchical solution presented earlier, there is cleaner separation in the K-Means plot. We will see how these clusters are separated by inspecting features within each group.

```{r}
# performs clustering on data matrix 
# number of clusters determined before
km_k <- 5


km_four <- kmeans(x = nba_feat_sc,
            centers = km_k,
            nstart = 100,
            algorithm = 'Hartigan-Wong')

nba$km_labs_four <- factor(km_four$cluster)
```


```{r echo=FALSE}
# swap labels 
# those who are 4 will become 1 (subpar)
# those who are 3 will necome 2 (below average)

nba$km_cluster <- factor(
  ifelse(nba$km_labs_four == 4, 1,
         ifelse(nba$km_labs_four == 3, 2, 
                ifelse(nba$km_labs_four == 2, 3, 
                       ifelse(nba$km_labs_four == 1, 4, 
                              5)))))

```

```{r echo=FALSE}
# rearrange in appropriate order

nba <- nba[, c(18, 1, 2, 40, 35, 3:6, 17, 7:10, 19, 20, 11:16, 21, 34, 22:33, 36:38 )]

```


```{r, echo=FALSE}
# plot k-means clusters in PC space
# Labels: Players who played more than 36 min per game or less than 3 min per game
km_labels <- ifelse(nba$MP_pg >= 36 | nba$MP_pg <= 3, 
                         as.character(nba$Player), '' )
 
# Cluster Plot 
plot_pca(
  km_four,
  data = nba,
  frame = TRUE,
  colour = 'km_cluster',
  title = paste0(
    km_k,
    ' Clusters (K-means)'
    )
  ) +
  theme_davide() + 
  theme(
    legend.margin = margin(4, 4, 4, 4), 
    legend.background = element_rect(fill = 'grey98'), 
    legend.position = 'bottom',
    plot.title = element_text(margin = margin(b=15, t = 10), size=30, hjust = 0.4, family = 'Proxima Nova', face = 'bold'),
    plot.margin = margin(5, 0, 5, 0)
    )



# ggsave('/Users/davidetissino/Desktop/Clusters.png', dpi = 'retina', width = 11, height = 9)

```

One explanation for the imbalanced number of players across clusters is that player skillset level is inherently imbalanced. This imbalance is reflected in the univariate plots, where most densities were positively skewed. This suggests that there a few players whose statistics significantly exceed those of the average player. This is also reflected in the average statistics by cluster shown below.

```{r echo=FALSE}
# Group by cluster and count total players in each 
cluster_tot <- nba %>% 
  group_by(
    km_cluster
  ) %>% 
  summarise(
    Count = n()
  ) %>% 
  ungroup()


# Plot pie with players / total 
ggplot(cluster_tot, aes(x = "", y = Count, fill = km_cluster)) +
  ggtitle('Players in each Cluster ') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = Count
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = 6) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  )

# ggsave('/Users/davidetissino/Desktop/Players_in_clusters.png', dpi = 'retina', width = 9, height = 9)


```

Players in each position within each cluster

```{r echo=FALSE}
# Group by cluster (km_cluster) and position (Pos), then count players
cluster_summary <- nba %>%
  group_by(km_cluster, Pos) %>%  # Group by cluster and position
  summarise(Count = n()) %>%  # Count number of players in each group
  ungroup() %>%  # Ungroup for the next summarization
  group_by(km_cluster) %>%  # Group by cluster again to add totals
  mutate(Total_Players = sum(Count))  # Calculate total players in each cluster

cluster_summary %>% 
  kbl(
    caption = 'K-Means Player Distribution'
  ) %>% 
  kable_styling(
    latex_options = c('striped')
  )

```

Players averages within each cluster 

```{r echo=FALSE}
# averages by cluster
nba_km_avg <- nba %>%
  # select(km_cluster, MP_pg, PTS_pg, TRB_pg,
  #        AST_pg, BLK_pg, STL_pg, VORP, PER, RPM, ) %>%
  group_by(km_cluster) %>% 
  summarise_all(list(mean))


# Average Stats by each Cluster
nba_km_avg %>%
  kbl(
    aption = 'Average Stats by Cluster',
      digits = 2
  ) %>% 
  kable_styling(
    latex_options = c('striped', 'scale_down')
  )

```

Player averages and total of cluster 

```{r echo=FALSE}

nba_km <- merge(
  cluster_tot, 
  nba_km_avg,
  by = 'km_cluster'
)



nba_km %>%
  kbl(
    aption = 'Stats by Cluster',
      digits = 2
  ) %>% 
  kable_styling(
    latex_options = c('striped', 'scale_down')
  )

```


```{r echo=FALSE}
# Division of Positions within Cluster
cluster_count <- nba_km[, c(1,2)]

cluster_count$pct <- round(cluster_count$Count / sum(cluster_count$Count), 3) * 100
  
# Plot positions within cluster
ggplot(cluster_count, aes(x = "", y = pct, fill = km_cluster)) +
  ggtitle('Clusters Composition') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = pct
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = ) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  )

# ggsave('/Users/davidetissino/Desktop/Cluster_Composition.png', dpi = 'retina', width = 9, height = 9)

```



## Final cluster selection
K-Means (four clusters) was the optimal solution. Comparing the HCL and KM cluster plots (per above) reveals the K-Means produces clearer separation of players based on overall skillsets. 

We validated this by comparing the clusters against advanced statistics. PER, VORP, and RPM are advanced statistics commonly used to assess general player performance. None of these statistics were used in the cluster modeling. 

```{r echo=FALSE}
# Plots advanced statistics for clusters
nba_km_avg %>%
  # Reorder km_labs_four factor levels
  select(km_cluster, VORP, PER, RPM) %>%
  pivot_longer(cols = c("VORP", "PER", "RPM")) %>%
  ggplot(aes(x = km_cluster, y = value, fill = km_cluster)) +
  geom_col() +
  facet_wrap(~name, scales = 'free') +
  labs(
    title = 'Overall Player Performance by Cluster',
    subtitle = 'Average Advanced Statistics',
    x = 'Cluster',
    y = 'Advanced Statistic Value'
  ) + 
  theme_davide() + 
  theme(
    plot.subtitle = element_text(margin = margin(b = 10)), 
    legend.position = 'none',
    strip.text = element_text(face = 'bold', size = 15)
  )


# ggsave('/Users/davidetissino/Desktop/Clusters_Adv_Stats.png', dpi = 'retina', width = 9, height = 9)



```



```{r echo=FALSE}
# Plots PTS, REB and AST for clusters
nba_km_avg %>%
  # Reorder km_labs_four factor levels
  select(km_cluster, PTS_pg, TRB_pg, AST_pg) %>%
  pivot_longer(cols = c("PTS_pg", "TRB_pg", "AST_pg")) %>%
  ggplot(aes(x = km_cluster, y = value, fill = km_cluster)) +
  geom_col() +
  facet_wrap(~name, scales = 'free') +
  labs(
    title = 'Overall Player Performance by Cluster',
    subtitle = 'Average Traditional Statistics',
    x = 'Cluster',
    y = 'Traditional Statistic Value'
  ) + 
  theme_davide() + 
  theme(
    plot.subtitle = element_text(margin = margin(b = 10)), 
    legend.position = 'none',
    strip.text = element_text(face = 'bold', size = 15)
  )

# ggsave('/Users/davidetissino/Desktop/Clusters_Trad_Stats.png', dpi = 'retina', width = 9, height = 9)



```

ANALYSIS OF EACH CLUSTER

STAR PLAYERS

```{r echo=FALSE}
## Specific plot for CLUSTER = 5, STARS

nba_stars <- nba_km_avg[5 ,]

nba_stars <- nba_stars %>%
  # Reorder km_labs_four factor levels
  select(km_cluster, FGA_pg, X3PA_pg, STL_pg, BLK_pg, AST_pg, TRB_pg, PER, BPM, VORP, PTS_pg) %>%
  pivot_longer(cols = c("FGA_pg", "X3PA_pg", "STL_pg", 'BLK_pg', 'AST_pg', 'TRB_pg', 'PER', 'BPM', 'VORP', 'PTS_pg'))

             

nba_stars$name <- factor(nba_stars$name, levels = c('PTS_pg', 'TRB_pg', 'AST_pg', 'STL_pg', 'BLK_pg', 'X3PA_pg', 'FGA_pg', 'RPM', 'VORP', 'PER'))  
               
nba_stars %>%
  ggplot(aes(x = name, y = value), fill = 'red') +
  geom_col() +
  geom_label(
    aes(label = round(value, 1)),
    fontface = 'bold', 
    color = 'black',
    fill = 'grey90',
    size = 7, 
    position = position_stack(vjust = 0.5)
  ) +
  labs(
    title = 'Overall Player Performance by Cluster',
    subtitle = 'Average Traditional and Advanced Statistics',
    x = 'Cluster 5: Stars',
    y = 'Statistic Value'
  ) + 
  theme_davide() + 
  theme(
    plot.subtitle = element_text(margin = margin(b = 10)), 
    legend.position = 'none',
    strip.text = element_text(face = 'bold', size = 15)
  ) +
  scale_fill_brewer(palette="Set1")




```

```{r echo=FALSE}
# Division of Positions within Cluster
cluster_summ_stars <- cluster_summary %>% 
  filter(
    km_cluster == 5
  )
  
# Plot positions within cluster
ggplot(cluster_summ_stars, aes(x = "", y = Count, fill = Pos)) +
  ggtitle('Cluster 5: All-Stars') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = paste0(Count, ' - ', Pos)
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = 10) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  )

# ggsave('/Users/davidetissino/Desktop/All-Stars_Positions.png', dpi = 'retina', width = 9, height = 9)

```

ROLE PLAYERS

```{r echo=FALSE}
## Specific plot for CLUSTER = 4, ROLE

nba_role <- nba_km_avg[4 ,]

nba_role <- nba_role %>%
  # Reorder km_labs_four factor levels
  select(km_cluster, FGA_pg, X3PA_pg, STL_pg, BLK_pg, AST_pg, TRB_pg, PER, BPM, VORP, PTS_pg) %>%
  pivot_longer(cols = c("FGA_pg", "X3PA_pg", "STL_pg", 'BLK_pg', 'AST_pg', 'TRB_pg', 'PER', 'BPM', 'VORP', 'PTS_pg'))

             
nba_role$name <- factor(nba_role$name, levels = c('PTS_pg', 'TRB_pg', 'AST_pg', 'STL_pg', 'BLK_pg', 'X3PA_pg', 'FGA_pg', 'RPM', 'VORP', 'PER'))  
               
nba_role %>%
  ggplot(aes(x = name, y = value, fill = km_cluster)) +
  geom_col() +
  geom_label(
    aes(label = round(value, 1)),
    fontface = 'bold', 
    color = 'black',
    fill = 'grey90',
    size = 7, 
    position = position_stack(vjust = 0.5)
  ) +
  labs(
    title = 'Overall Player Performance by Cluster',
    subtitle = 'Average Traditional and Advanced Statistics',
    x = 'Cluster 4: Rim-Centric Players',
    y = 'Statistic Value'
  ) + 
  theme_davide() + 
  theme(
    plot.subtitle = element_text(margin = margin(b = 10)), 
    legend.position = 'none',
    strip.text = element_text(face = 'bold', size = 15)
  ) 


```

```{r echo=FALSE}
# Division of Positions within Cluster
cluster_summ_role <- cluster_summary %>% 
  filter(
    km_cluster == 4
  )
  
# Plot positions within cluster
ggplot(cluster_summ_role, aes(x = "", y = Count, fill = Pos)) +
  ggtitle('Cluster 4: Rim-Centric Players') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = ifelse(
        Count == 1, '',  paste0(Count, ' - ', Pos)
      )
    ),
  position = position_stack(vjust = 0.5),
  color = 'white', 
  fontface = 'bold', 
  size = 10) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  ) 

# ggsave('/Users/davidetissino/Desktop/Rim-Centric_Positions.png', dpi = 'retina', width = 9, height = 9)


```

BENCH PLAYERS

```{r echo=FALSE}
## Specific plot for CLUSTER = 3, BENCH

nba_bench <- nba_km_avg[3 ,]

nba_bench <- nba_bench %>%
  # Reorder km_labs_four factor levels
  select(km_cluster, FGA_pg, X3PA_pg, STL_pg, BLK_pg, AST_pg, TRB_pg, PER, BPM, VORP, PTS_pg) %>%
  pivot_longer(cols = c("FGA_pg", "X3PA_pg", "STL_pg", 'BLK_pg', 'AST_pg', 'TRB_pg', 'PER', 'BPM', 'VORP', 'PTS_pg'))

             

nba_bench$name <- factor(nba_bench$name, levels = c('PTS_pg', 'TRB_pg', 'AST_pg', 'STL_pg', 'BLK_pg', 'X3PA_pg', 'FGA_pg', 'BPM', 'VORP', 'PER'))  
               
nba_bench %>%
  ggplot(aes(x = name, y = value, fill = km_cluster)) +
  geom_col() +
  geom_label(
    aes(label = round(value, 1)),
    fontface = 'bold', 
    color = 'black',
    fill = 'grey90',
    size = 7, 
    position = position_stack(vjust = 0.5)
  ) + 
  labs(
    title = 'Overall Player Performance by Cluster',
    subtitle = 'Average Traditional and Advanced Statistics',
    x = 'Cluster 3: Second Options',
    y = 'Statistic Value'
  ) + 
  theme_davide() + 
  theme(
    plot.subtitle = element_text(margin = margin(b = 10)), 
    legend.position = 'none',
    strip.text = element_text(face = 'bold', size = 15)
  ) 

# ggsave('/Users/davidetissino/Desktop/Bench_Positions.png', dpi = 'retina', width = 9, height = 9)

```


```{r echo=FALSE}

# Division of Positions within Cluster
cluster_summ_bench <- cluster_summary %>% 
  filter(
    km_cluster == 3
  )
  
# Plot positions within cluster
ggplot(cluster_summ_bench, aes(x = "", y = Count, fill = Pos)) +
  ggtitle('Cluster 3: Second Options') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = ifelse(
        Pos == 'C', 
        '', 
        paste0(Count, ' - ', Pos)
      )
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = 10) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  ) 

# 
ggsave('/Users/davidetissino/Desktop/Second_Options_Positions.png', dpi = 'retina', width = 9, height = 9)


```

UNDERPERFORMING PLAYERS

```{r echo=FALSE}
## Specific plot for CLUSTER = 2, UNDERPERFORMING

nba_below_avg <- nba_km_avg[2 ,]

nba_below_avg <- nba_below_avg %>%
  # Reorder km_labs_four factor levels
  select(km_cluster, FGA_pg, X3PA_pg, STL_pg, BLK_pg, AST_pg, TRB_pg, PER, BPM, VORP, PTS_pg) %>%
  pivot_longer(cols = c("FGA_pg", "X3PA_pg", "STL_pg", 'BLK_pg', 'AST_pg', 'TRB_pg', 'PER', 'BPM', 'VORP', 'PTS_pg'))

             

nba_below_avg$name <- factor(nba_below_avg$name, levels = c('PTS_pg', 'TRB_pg', 'AST_pg', 'STL_pg', 'BLK_pg', 'X3PA_pg', 'FGA_pg', 'BPM', 'VORP', 'PER'))  
               
nba_below_avg %>%
  ggplot(aes(x = name, y = value, fill = km_cluster)) +
  geom_col() +
  geom_label(
    aes(label = round(value, 1)),
    fontface = 'bold', 
    color = 'black',
    fill = 'grey90',
    size = 7, 
    position = position_stack(vjust = 0.5)
  ) + 
  labs(
    title = 'Overall Player Performance by Cluster',
    subtitle = 'Average Traditional and Advanced Statistics',
    x = 'Cluster 2: Underperforming Players',
    y = 'Statistic Value'
  ) + 
  theme_davide() + 
  theme(
    plot.subtitle = element_text(margin = margin(b = 10)), 
    legend.position = 'none',
    strip.text = element_text(face = 'bold', size = 15)
  ) 

```

```{r echo=FALSE}
# Division of Positions within Cluster
cluster_summ_below <- cluster_summary %>% 
  filter(
    km_cluster == 2
  )
  
# Plot positions within cluster
ggplot(cluster_summ_below, aes(x = "", y = Count, fill = Pos)) +
  ggtitle('Cluster 2: Underperforming Players') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = paste0(Count, ' - ', Pos)
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = 6) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  ) 

# ggsave('/Users/davidetissino/Desktop/Underperforming_Positions.png', dpi = 'retina', width = 9, height = 9)


```

LOW-TIER PLAYERS

```{r echo=FALSE}
## Specific plot for CLUSTER = 1, LOW-TIER

nba_subpar <- nba_km_avg[1 ,]

nba_subpar <- nba_subpar %>%
  # Reorder km_labs_four factor levels
  select(km_cluster, FGA_pg, X3PA_pg, STL_pg, BLK_pg, AST_pg, TRB_pg, PER, BPM, VORP, PTS_pg) %>%
  pivot_longer(cols = c("FGA_pg", "X3PA_pg", "STL_pg", 'BLK_pg', 'AST_pg', 'TRB_pg', 'PER', 'BPM', 'VORP', 'PTS_pg'))

             

nba_subpar$name <- factor(nba_subpar$name, levels = c('PTS_pg', 'TRB_pg', 'AST_pg', 'STL_pg', 'BLK_pg', 'X3PA_pg', 'FGA_pg', 'BPM', 'VORP', 'PER'))  
               
nba_subpar %>%
  ggplot(aes(x = name, y = value, fill = km_cluster)) +
  geom_col() +
  geom_label(
    aes(label = round(value, 1)),
    fontface = 'bold', 
    color = 'black',
    fill = 'grey90',
    size = 7, 
    position = position_stack(vjust = 0.5)
  ) + 
  labs(
    title = 'Overall Player Performance by Cluster',
    subtitle = 'Average Traditional and Advanced Statistics',
    x = 'Cluster 1: Low-Tier Players',
    y = 'Statistic Value'
  ) + 
  theme_davide() + 
  theme(
    plot.subtitle = element_text(margin = margin(b = 10)), 
    legend.position = 'none',
    strip.text = element_text(face = 'bold', size = 15)
  ) 

```

```{r echo=FALSE}
# Division of Positions within Cluster
cluster_summ_subpar <- cluster_summary %>% 
  filter(
    km_cluster == 1
  )
  
# Plot positions within cluster
ggplot(cluster_summ_subpar, aes(x = "", y = Count, fill = Pos)) +
  ggtitle('Cluster 1: Low-Tier Players') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = paste0(Count, ' - ', Pos)
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = 6) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  ) 


```


```{r echo=FALSE}
par(mfrow = c(3,2))

ggplot(cluster_summ_subpar, aes(x = "", y = Count, fill = Pos)) +
  ggtitle('Cluster 1: Low-Tier Players') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = paste0(Count, ' - ', Pos)
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = 6) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  ) 


ggplot(cluster_summ_below, aes(x = "", y = Count, fill = Pos)) +
  ggtitle('Cluster 2: Underperforming Players') +
  geom_col(color = "black") +
  geom_text(
    aes(
      label = paste0(Count, ' - ', Pos)
    ),
    position = position_stack(vjust = 0.5),
    color = 'white', 
    fontface = 'bold', 
    size = 6) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("#BE2A3E", "#EC754A",
                               "#FFC125", "#3C8D53", '#4876FF'))+
  theme_davide() +
  theme(
    axis.title.y = element_blank(), 
    axis.title.x = element_blank(),
    legend.position = 'bottom', 
    legend.direction = 'horizontal', 
    legend.margin = margin(3, 3, 3, 3), 
    plot.margin = margin(t = 15, b = 25), 
    legend.text = element_text(size = 13)
  ) 






```

